---
title: "Example Usage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example-usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tabtibble)
library(dplyr)
```

The `tabtibble` package is typically used to create a list of tables for
reporting, often in `quarto` or `rmarkdown`.

A simple example is to create a list of tables with each number of cylinders
grouped together.

```{r example-setup}
d_tab <-
  mtcars %>%
  tidyr::nest(table = !"cyl") %>%
  dplyr::mutate(
    caption = glue::glue("Cars with {cyl} cylinders")
  ) %>%
  new_tab_tibble()
```

Then, print it! By default (the `tabtibble.knitr.auto.asis` option, `TRUE`),
no `results='asis'` chunk option is needed -- the same mechanism pander uses
for its `knitr.auto.asis` option, returning the rendered markdown as
`knitr::asis_output()` so knitr inserts it directly.

```{r print}
knit_print(d_tab)
```

Set `options(tabtibble.knitr.auto.asis = FALSE)` to go back to writing
directly with `cat()` instead, which does require `results='asis'` on the
chunk (and warns when it is missing).

## Quarto + Typst reports

Each table is wrapped in a fenced Div with a Quarto label (`#tbl-<label>`),
so it can be cross-referenced with `@tbl-<label>` and, when the report is
rendered to a Typst PDF, appears in a Typst list of tables. Add a label
column when you want to choose the label yourself; otherwise a stable one is
derived from the caption, so re-rendering the same report reproduces the
same labels:

```{r label-example}
d_tab_labelled <-
  mtcars %>%
  tidyr::nest(table = !"cyl") %>%
  dplyr::mutate(
    caption = glue::glue("Cars with {cyl} cylinders"),
    label = glue::glue("cyl-{cyl}")
  ) %>%
  new_tab_tibble()
d_tab_labelled$label
```

In the Typst document, add a list of tables near the top with:

````
```{=typst}
#outline(title: [List of Tables], target: figure.where(kind: "quarto-float-tbl"))
```
````

(`"quarto-float-tbl"` is Quarto's own kind string for a crossreferenceable
table from computational output; the native Typst `table` kind is for tables
written directly in Typst source.)

## Rendering backends

By default, each table is rendered as a Markdown (pipe) table via
`knitr::kable()`, so `tabtibble` itself needs no extra table-formatting
package. Set the `tabtibble.backend` option to `"tinytable"`, `"gt"`, or
`"flextable"` to render plain data.frame tables with that package instead
(each package is only required when you select its backend):

```{r backend-example}
options(tabtibble.backend = "tinytable")
```

A table you already built with `gt`, `tinytable`, `flextable`, or `table1`
is printed exactly as you built it, using that package's own printing
method, regardless of `tabtibble.backend`.

Caption and cell text containing Typst-significant characters (`* _ # @ $ <
> \` [ ]` or a backslash) render literally; `tabtibble` escapes them once,
for every backend, so you never need to escape them yourself.

## Topic-grouped listings (`"typst"` backend)

The `"typst"` backend reproduces what LaTeX's TopicLongTable gave: a listing
sorted by one or more "topic" columns (for example subject, then analyte)
that prints each topic value once, as a group header above its rows, instead
of repeating it on every row. When the group runs onto the next page, the
column header and the current topic repeat at the top of that page, so a
reader never sees a page of data with no idea which subject it belongs to.
Declare the topic columns with `topic_cols`; the underlying table must
already be sorted by them:

```{r topic-example}
conc <-
  tibble::tibble(
    subject = rep(c("S001", "S002"), each = 3),
    time    = rep(0:2, times = 2),
    conc    = c(0, 5.1, 4.3, 0, 6.0, 5.2)
  )

d_tab_topic <-
  tibble::tibble(
    table      = list(conc),
    caption    = "Concentrations by subject",
    label      = "conc-by-subject",
    topic_cols = list("subject")
  ) %>%
  new_tab_tibble()

options(tabtibble.backend = "typst")
```

```{r topic-print, results='asis'}
knit_print(d_tab_topic)
```

Without `topic_cols`, the `"typst"` backend is an ordinary table whose
column header repeats on every page. The other backends ignore
`topic_cols` gracefully: the named columns stay as ordinary columns, so
their values just repeat on every row.

Column alignment defaults to right for numeric columns and left for
everything else; override it with a `tabtibble_align` attribute (a named
character vector of `"left"`, `"right"`, or `"center"`) set on the table
before calling `new_tab_tibble()`:

```{r topic-align-example}
attr(conc, "tabtibble_align") <- c(subject = "left", time = "center", conc = "right")
d_tab_topic_aligned <-
  tibble::tibble(table = list(conc), caption = "Aligned columns", topic_cols = list("subject")) %>%
  new_tab_tibble()
```
