---
title: "Example Usage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example-usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tabtibble)
library(dplyr)
```

The `tabtibble` package is typically used to create a list of tables for
reporting, often in `quarto` or `rmarkdown`.

A simple example is to create a list of tables with each number of cylinders
grouped together.

```{r example-setup}
d_tab <-
  mtcars %>%
  tidyr::nest(table = !"cyl") %>%
  dplyr::mutate(
    caption = glue::glue("Cars with {cyl} cylinders")
  ) %>%
  new_tab_tibble()
```

Then, print it! You must set `results='asis'` for the chunk printing your
`tab_tibble`.

```{r print, results='asis'}
knit_print(d_tab)
```

## Quarto + Typst reports

Each table is wrapped in a fenced Div with a Quarto label (`#tbl-<label>`),
so it can be cross-referenced with `@tbl-<label>` and, when the report is
rendered to a Typst PDF, appears in a Typst list of tables. Add a label
column when you want to choose the label yourself; otherwise a stable one is
derived from the caption, so re-rendering the same report reproduces the
same labels:

```{r label-example}
d_tab_labelled <-
  mtcars %>%
  tidyr::nest(table = !"cyl") %>%
  dplyr::mutate(
    caption = glue::glue("Cars with {cyl} cylinders"),
    label = glue::glue("cyl-{cyl}")
  ) %>%
  new_tab_tibble()
d_tab_labelled$label
```

In the Typst document, add a list of tables near the top with:

````
```{=typst}
#outline(title: [List of Tables], target: figure.where(kind: "quarto-float-tbl"))
```
````

(`"quarto-float-tbl"` is Quarto's own kind string for a crossreferenceable
table from computational output; the native Typst `table` kind is for tables
written directly in Typst source.)

## Rendering backends

By default, each table is rendered as a Markdown (pipe) table via
`knitr::kable()`, so `tabtibble` itself needs no extra table-formatting
package. Set the `tabtibble.backend` option to `"tinytable"`, `"gt"`, or
`"flextable"` to render plain data.frame tables with that package instead
(each package is only required when you select its backend):

```{r backend-example}
options(tabtibble.backend = "tinytable")
```

A table you already built with `gt`, `tinytable`, `flextable`, or `table1`
is printed exactly as you built it, using that package's own printing
method, regardless of `tabtibble.backend`.

Caption and cell text containing Typst-significant characters (`* _ # @ $ <
> \` [ ]` or a backslash) render literally; `tabtibble` escapes them once,
for every backend, so you never need to escape them yourself.
